---
title: "Lantz, Brett - Machine Learning with R (4th ed.), Chapter 8: Finding Patterns - Market Basket Analysis Using Association Rules"
author: 'Original Code: Brett Lantz | Modifications: Antti Rask'
date: "2023-07-23"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading Packages

```{r}
library(arules)
library(arulesViz)
library(conflicted)
library(tidyverse)
```

## Exploring and preparing the data

### Load the grocery data into a sparse matrix

```{r}
groceries <- read.transactions(
    "data/groceries.csv",
    sep = ","
)
```

```{r}
groceries %>% 
    summary()
```

### Look at the first five transactions

```{r}
groceries %>%
    head(5) %>%
    inspect()
```

### Examine the frequency of items

```{r}
groceries[, 1:3] %>%
    itemFrequency()
```

### Plot the frequency of items

Have to adjust the margins or there will be an error

```{r}
par(mar = c(1, 1, 1, 1))
```

```{r}
groceries %>%
    itemFrequencyPlot(support = 0.1)
```

```{r}
groceries %>%
    itemFrequencyPlot(topN = 20)
```

### A visualization of the sparse matrix for the first five transactions

```{r}
groceries[1:5] %>%
    image()
```

### Visualization of a random sample of 100 transactions

```{r}
sample(groceries, 100) %>%
    image()
```

## Training a model on the data

### Default settings result in zero rules learned

```{r}
groceries %>%
    apriori()
```

### Set better support and confidence levels to learn more rules

```{r}
groceryrules <- apriori(
    groceries,
    parameter = list(
        support = 0.006,
        confidence = 0.25,
        minlen = 2
    )
)

groceryrules
```

## Evaluating model performance

### Summary of grocery association rules

```{r}
groceryrules %>% 
    summary()
```

### Look at the first three rules

```{r}
groceryrules[1:3] %>%
    inspect()
```

### Visualize the rules

```{r}
groceryrules %>% plot(
    method = "graph",
    limit  = 10
)
```

## Improving model performance

### Sorting grocery rules by lift

```{r}
sort(groceryrules, by = "lift")[1:5] %>%
    inspect()
```

### Finding subsets of rules containing any berry items

```{r}
berryrules <- groceryrules %>%
    subset(items %in% "berries")
```

```{r}
berryrules %>%
    inspect()
```

### Converting the rule set to a tibble

```{r}
groceryrules_tbl <- groceryrules %>%
    as("data.frame") %>%
    as_tibble()
```

```{r}
groceryrules_tbl
```
