---
title: "Lantz, Brett - Machine Learning with R (4th ed.), Chapter 9: Finding Groups of Data - Clustering with k-means"
author: 'Original Code: Brett Lantz | Modifications: Antti Rask'
date: "2023-07-23"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading Packages

```{r}
library(conflicted)
library(janitor)
library(tidyclust)
library(tidymodels)
library(tidyverse)
```

## Exploring and preparing the data

```{r}
teens_tbl <- read_csv("data/snsdata.csv") %>%
    mutate(across(where(is.character), as.factor))
```

```{r}
teens_tbl %>%
    glimpse()
```

### Look at missing data for age variable

```{r}
teens_tbl %>%
    pull(age) %>%
    summary()
```

### Eliminate age outliers

```{r}
teens_only_tbl <- teens_tbl %>%
    mutate(
        age = case_when(
            between(age, 13, 20) ~ age,
            .default             = NA_real_
        )
    )
```

```{r}
teens_only_tbl %>%
    pull(age) %>%
    summary()
```

### Look at missing data for female variable

```{r}
teens_only_tbl %>%
    tabyl(gender) %>%
    adorn_pct_formatting(digits = 1)
```

### Reassign missing gender values to "unknown" and change into factors

```{r}
teens_gendered_tbl <- teens_only_tbl %>%
    mutate(
        gender = case_when(
            gender %in% c("F", "M") ~ gender,
            .default                = "gender_unknown"
        ) %>%
            as.factor()
    )
```

### Check our recoding work

```{r}
teens_gendered_tbl %>%
    tabyl(gender) %>%
    adorn_pct_formatting(digits = 1) %>%
    arrange(desc(percent))
```

### Finding the mean age by cohort

Doesn't work because of the NAs

```{r}
teens_gendered_tbl %>%
    pull(age) %>%
    mean()
```

Works thanks to the na.rm = TRUE

```{r}
teens_gendered_tbl %>%
    pull(age) %>%
    mean(na.rm = TRUE)
```

### Age by cohort

```{r}
teens_gendered_tbl %>%
    summarize(
        age = mean(age, na.rm = TRUE),
        .by = gradyear
    )
```

### Create a vector with the average age for each gradyear, repeated by person

```{r}
average_age_by_gradyear <- teens_gendered_tbl %>%
    mutate(
        ave_age = mean(age, na.rm = TRUE),
        .by     = gradyear
    ) %>%
    pull()
```

### Impute the missing age values with the average age by gradyear

```{r}
teens_imputed_tbl <- teens_gendered_tbl %>%
    mutate(
        age = case_when(
            !is.na(age) ~ age,
            .default    = average_age_by_gradyear
        )
    )
```

### Check the summary results to ensure missing values are eliminated

```{r}
teens_imputed_tbl %>%
    pull(age) %>%
    summary()
```

## Creating the recipe

### Apply normalization to entire data frame

```{r}
recipe_obj <- recipe(teens_imputed_tbl) %>%
    step_normalize(
        all_numeric(),
        -c(gradyear, age, friends)
    ) %>%
    step_dummy(
        gender,
        keep_original_cols = TRUE,
        one_hot            = TRUE
    )

recipe_obj
```

```{r}
recipe_baked_tbl <- recipe_obj %>%
    prep() %>%
    bake(new_data = NULL)

recipe_baked_tbl
```

## Training a model on the data

### Create a z-score standardized data frame for easier interpretation

```{r}
interests_tbl <- recipe_baked_tbl %>%
    select(5:40)
```

### Compare the data before and after the transformation

Before

```{r}
teens_imputed_tbl %>%
    pull(basketball) %>%
    summary()
```

After

```{r}
interests_tbl %>%
    pull(basketball) %>%
    summary()
```

#### Create model specification

```{r}
model_spec_kmeans <- k_means(
    engine       = "stats",
    num_clusters = 5
)

model_spec_kmeans
```

```{r}
set.seed(2345)

model_fit_kmeans <- model_spec_kmeans %>%
  fit(
      ~ .,
      interests_tbl
  )

model_fit_kmeans
```

## Evaluating model performance

### Look at the single-row summary

```{r}
model_fit_kmeans %>%
    glance()
```

### Look at the size and the centers of the clusters

```{r}
model_fit_kmeans %>%
    tidy() %>%
    select(cluster, size, withinss, everything())
```

## Improving model performance

### Apply the cluster IDs to the original data frame

```{r}
teens_and_clusters <- augment(model_fit_kmeans, recipe_baked_tbl)
teens_and_clusters
```

### Look at the first five records

```{r}
teens_and_clusters %>%
    select(.pred_cluster, gender, age, friends) %>%
    slice_head(n = 5)
```

### Mean age by cluster

```{r}
teens_and_clusters %>%
    summarize(
        age = mean(age),
        .by = .pred_cluster
    )
```

### Proportion of females by cluster

```{r}
teens_and_clusters %>%
    summarize(
        gender_F = mean(gender_F),
        .by      = .pred_cluster 
    )
```

### Mean number of friends by cluster

```{r}
teens_and_clusters %>%
    summarize(
        friends = mean(friends),
        .by     = .pred_cluster
    )
```

## K-means clustering with tidy data principles

### Exploratory clustering

```{r}
kclusts <-
    tibble(k = 1:9) %>%
    mutate(
        kclust    = map(k, ~kmeans(interests_tbl, .x)),
        tidied    = map(kclust, tidy),
        glanced   = map(kclust, glance),
        augmented = map(kclust, augment, recipe_baked_tbl)
    )
```

### Create three separate datasets

```{r}
clusters <- kclusts %>%
    unnest(cols = c(tidied))

assignments <- kclusts %>%
    unnest(cols = c(augmented))

clusterings <- kclusts %>%
    unnest(cols = c(glanced))
```

### Plot the original points

```{r}
p1 <- assignments %>%
    ggplot(aes(sports, music)) +
    geom_point(aes(color = .cluster), alpha = 0.5) +
    facet_wrap(vars(k)) +
    theme_bw()

p1
```

```{r}
p2 <- p1 +
    geom_point(data = clusters, size = 5, shape = "x")

p2
```

### Plot the total within sum of squares (tot.withinss)

```{r}
clusterings %>%
    ggplot(aes(k, tot.withinss)) +
    geom_line() +
    geom_point() +
    theme_bw() +
    scale_x_continuous(breaks = seq(1, 9, by = 1)) +
    labs(
        x = "k",
        y = "Total Within Sum of Squares"
    )
```
