---
title: "Lantz, Brett - Machine Learning with R (4th ed.), Chapter 2: Managing and Understanding Data"
author: "Original Code: Brett Lantz | Modifications: Antti Rask"
date: "2023-07-23"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 2 Managing and Understanding Data

## Load Packages

```{r}
library(conflicted) # An Alternative Conflict Resolution Strategy
    conflicts_prefer(dplyr::filter)
    conflicts_prefer(janitor::chisq.test)
library(crosstable) # Crosstables for Descriptive Analyses
library(janitor)    # Simple Tools for Examining and Cleaning Dirty Data
library(tidyverse)  # Easily Install and Load the 'Tidyverse'
```

## Data

```{r}
# create tibble of data for three medical patients
patients_tbl <- tibble(
    subject_name = c("John Doe", "Jane Doe", "Steve Graves"),
    temperature  = c(98.1, 98.6, 101.4),
    flu_status   = c(FALSE, FALSE, TRUE),
    gender       = factor(c("MALE", "FEMALE", "MALE")),
    blood        = factor(
        c("O", "AB", "A"),
        levels = c("A", "B", "AB", "O")
    ),
    symptoms     = factor(
        c("SEVERE", "MILD", "MODERATE"),
        levels = c("MILD", "MODERATE", "SEVERE"),
        ordered = TRUE
    )
)

patients_tbl
```

## R data structures

### Vectors -> Tibble

```{r}
# access the second body temperature
patients_tbl %>%
    slice(2) %>%
    pull(temperature)
```

```{r}
# access body temperature for the second and third patient
patients_tbl %>%
    slice(2:3) %>%
    pull(temperature)
```

```{r}
# access body temperature for all but the second patient
patients_tbl %>%
    slice(-2) %>%
    pull(temperature)
```

```{r}
# Use a vector to indicate whether to include item
patients_tbl %>%
    filter(c(TRUE, TRUE, FALSE)) %>%
    pull(temperature)
```

```{r}
# filter rows by conditions
patients_tbl %>%
    filter(temperature > 100) %>% 
    pull(subject_name)
```

### Factors

```{r}
# check gender factors
patients_tbl %>%
    pull(gender)
```

```{r}
# check blood type factors
patients_tbl %>% 
    pull(blood)
```

```{r}
# check symptom factors
patients_tbl %>%
    pull(symptoms)
```

```{r}
# check for symptoms greater than moderate
patients_tbl %>%
    mutate(symptoms_severe = symptoms > "MODERATE") %>% 
    pull(symptoms_severe)
```

### Lists/Tibble

```{r}
# display information for a patient
patients_tbl %>%
    slice(1)
```

```{r}
# convert the tibble to a list
patient_1_list <- patients_tbl %>%
    slice(1) %>% 
    as.list()

patient_1_list
```

#### Methods for Accessing a List

```{r}
# get a single list value by position (returns a sub-list)
patient_1_list[2]
```

```{r}
# get a single list value by position (returns a numeric vector)
patient_1_list %>% 
    pluck(2)
```

```{r}
# get a single list value by name
patient_1_list %>% 
    pluck("temperature")
```

```{r}
# get several list items by specifying a vector of names
patient_1_list[c("temperature", "flu_status")]
```
#### Access a List Like a Vector

```{r}
# get values 2 and 3
patient_1_list[2:3]
```

### Tibble

#### Accessing a tibble

```{r}
# display the data tibble
patients_tbl
```

```{r}
# get a single column
patients_tbl %>%
    pull(subject_name)
```

```{r}
# get several columns by specifying a vector of names
patients_tbl %>% 
    select(c("temperature", "flu_status"))
```

```{r}
# this is the same as above, extracting temperature and flu_status
patients_tbl %>% 
    select(2:3)
```

```{r}
# accessing by row and column
patients_tbl %>% 
    slice(c(1, 3)) %>%
    select(c(2, 4))
```

#### Extract All Rows or Columns

```{r}
# column 1, all rows
patients_tbl %>% 
    pull(1)
```

```{r}
# row 1, all columns
patients_tbl %>%
    slice(1)
```

```{r}
# all rows and all columns
patients_tbl
```

```{r}
# the following are equivalent
patients_tbl %>% 
    slice(c(1, 3)) %>%
    select(c(temperature, gender))

patients_tbl %>%
    slice(-2) %>%
    select(c(-1, -3, -5, -6))
```

```{r}
# creating a Celsius temperature column, then comparing before and after
patients_tbl %>%
    mutate(temp_c = (temperature - 32) * (5 / 9)) %>%
    select(temperature, temp_c)
```

### Matrixes

```{r}
# create a 2x2 matrix
matrix(c(1, 2, 3, 4), nrow = 2)
```

```{r}
# equivalent to the above
matrix(c(1, 2, 3, 4), ncol = 2)
```

```{r}
# create a 2x3 matrix
matrix(c(1, 2, 3, 4, 5, 6), nrow = 2)
```

```{r}
# create a 3x2 matrix
m <- matrix(c(1, 2, 3, 4, 5, 6), ncol = 2)
m
```

```{r}
# extract values from matrixes
m[1, 1]
m[3, 2]
```

```{r}
# extract rows
m[1, ]
```

```{r}
# extract columns
m[, 1]
```

## Managing data with R

#### Saving, Loading, and Removing R Data Structures

```{r}
# show all data structures in memory
ls()
```

```{r}
# remove the m and patient_1_list objects
rm(m, patient_1_list)
ls()
```

```{r}
# remove all (remaining) objects
rm(list = ls())
ls()
```

## Importing and saving datasets from CSV files

```{r}
# reading a CSV file
patients_tbl <- read_csv("data/pt_data.csv")
```

```{r}
# reading a CSV file and converting all character columns to factors
patients_factors_tbl <- read_csv(
    "data/pt_data.csv",
    col_types      = "cdlff",
    show_col_types = TRUE
)
```

```{r}
# alternatively reading a CSV file and converting all character columns to factors
patients_factors_tbl <- read_csv("data/pt_data.csv") %>%
    mutate(across(where(is.character), as.factor))

patients_factors_tbl
```

## Exploring and understanding data

### Data Exploration Example Using Used Car Data

```{r}
usedcars <- read_csv("data/usedcars.csv")
```

```{r}
# get structure of used car data
usedcars %>%
    str()
```

### Exploring numeric variables

```{r}
# summarize numeric variables
usedcars %>% 
    select(year, price, mileage) %>% 
    summary()
```

```{r}
# calculate the mean income
(36000 + 44000 + 56000) / 3

c(36000, 44000, 56000) %>%
    mean()
```

```{r}
# the median income
c(36000, 44000, 56000) %>%
    median()
```

```{r}
# the min/max of used car prices
usedcars %>%
    pull(price) %>% 
    range()
```

```{r}
# the difference of the range
usedcars %>% 
    pull(price) %>%
    range() %>%
    diff()
```

```{r}
# IQR for used car prices
usedcars %>% 
    pull(price) %>%
    IQR()
```

```{r}
# use quantile to calculate five-number summary
usedcars %>%
    pull(price) %>%
    quantile()
```

```{r}
# the 99th percentile
usedcars %>% 
    pull(price) %>% 
    quantile(
        probs = c(0.01, 0.99)
    )
```

```{r}
# quintiles
usedcars %>% 
    pull(price) %>%
    quantile(
        seq(from = 0, to = 1, by = 0.20)
    )
```

```{r}
# boxplot of used car prices and mileage
usedcars %>%
    ggplot(aes(y = price)) +
    geom_boxplot() +
    labs(
        title = "Boxplot of Used Car Prices",
        y     = "Price ($)"
    ) +
    theme_classic()

usedcars %>%
    ggplot(aes(y = mileage)) +
    geom_boxplot() +
    labs(
        title = "Boxplot of Used Car Mileage",
        y     = "Odometer (mi.)"
    ) +
    theme_classic()
```

```{r}
# histograms of used car prices and mileage
usedcars %>%
    ggplot(aes(price)) +
    geom_histogram(
        bins  = 10,
        color = "black",
        fill  = "white"
    ) +
    labs(
        title = "Histogram of Used Car Prices",
        x     = "Price ($)"
    ) +
    theme_classic()

usedcars %>%
    ggplot(aes(mileage)) +
    geom_histogram(
        bins  = 8,
        color = "black",
        fill  = "white"
    ) +
    labs(
        title = "Histogram of Used Car Mileage",
        x     = "Odometer (mi.)"
    ) +
    theme_classic()
```

```{r}
# variance and standard deviation of the used car data
usedcars %>%
    pull(price) %>% 
    var()
```

```{r}
usedcars %>%
    pull(price) %>%
    sd()
```

```{r}
usedcars %>%
    pull(mileage) %>%
    var()
```

```{r}
usedcars %>%
    pull(mileage) %>%
    sd()
```

### Exploring categorical features

```{r}
# one-way tables for the used car data
usedcars %>% 
    tabyl(year) %>%
    adorn_pct_formatting(digits = 1)
```

```{r}
usedcars %>% 
    tabyl(model) %>%
    adorn_pct_formatting(digits = 1)
```

```{r}
usedcars %>% 
    tabyl(color) %>%
    adorn_pct_formatting(digits = 1)
```

```{r}
# you can also have a three-way table of the same
usedcars %>% 
    tabyl(year, model, color)
```

### Exploring relationships between variables

```{r}
# scatterplot of price vs. mileage
usedcars %>%
    ggplot(aes(mileage, price)) +
    geom_point() +
    labs(
        title = "Scatterplot of Price vs. Mileage",
        x     = "Used Car Odometer (mi.)",
        y     = "Used Car Price ($)"
    ) +
    theme_classic()
```

```{r}
# new variable indicating conservative colors
usedcars_conservative <- usedcars %>%
    mutate(conservative = color %in% c("Black", "Gray", "Silver", "White"))
```

```{r}
# checking our variable
usedcars_conservative %>% 
    tabyl(conservative) %>%
    adorn_pct_formatting(digits = 0)
```

```{r}
# crosstab of model by conservative 
usedcars_conservative %>%
    crosstable(
        cols            = model,
        by              = conservative,
        label           = FALSE,
        total           = "both",
        percent_pattern = "{n} ({p_row}/{p_col}/{p_tot})",
        percent_digits  = 1
    ) %>%
    as_flextable(compact = TRUE)
```

```{r}
# compute the chi-square test
usedcars_conservative %>%
    tabyl(model, conservative) %>%
    chisq.test()
```

