---
title: "Lantz, Brett - Machine Learning with R (4th ed.), Chapter 4: Probabilistic Learning - Classification Using Naive Bayes"
author: 'Original Code: Brett Lantz | Modifications: Antti Rask'
date: "2023-07-23"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load packages

```{r}
library(conflicted)  # An Alternative Conflict Resolution Strategy
conflict_prefer("filter", "dplyr", "stats")
library(crosstable)  # Crosstables for Descriptive Analyses
library(discrim)     # Model Wrappers for Discriminant Analysis
library(janitor)     # Simple Tools for Examining and Cleaning Dirty Data
library(SnowballC)   # Snowball Stemmers Based on the C 'libstemmer' UTF-8 Library
library(textrecipes) # Extra 'Recipes' for Text Processing
library(tidymodels)  # Easily Install and Load the 'Tidymodels' Packages
library(tidytext)    # Text Mining using 'dplyr', 'ggplot2', and Other Tidy Tools
library(tidyverse)   # Easily Install and Load the 'Tidyverse'
library(wordcloud2)  # Create Word Cloud by 'htmlwidget'
```

## Explore and prepare the data

### Read the sms data into the sms tibble, convert spam/ham to factor

```{r}
sms_tbl <- read_csv(
    "data/sms_spam.csv",
    col_types = "fc"
) %>%
    rename(.type = type)

sms_tbl
```

### Examine the structure of the sms data

```{r}
glimpse(sms_tbl)
```

### Examine the distribution of spam/ham

```{r}
sms_tbl %>%
    tabyl(.type) %>%
    adorn_pct_formatting(digits = 1)
```

### Build a corpus using the {tidytext} package instead of {tm}

This part inspired by a blog post by Julia Silge: https://www.tidyverse.org/blog/2020/11/tidymodels-sparse-support/

### Add row numbers

```{r}
sms_tbl_row_numbers <- sms_tbl %>% 
    mutate(line = row_number()
    )
```



### Manual preprocessing

```{r}
tidy_sms_tbl <- sms_tbl_row_numbers %>%
    
    mutate(
        
        # Make everything lowercase
        text = str_to_lower(text),
        
        # Remove numbers and punctuation,
        text = str_replace_all(text, "[[:punct:][:digit:]]+", " ")
        
    ) %>%
    
    # Unnest tokens
    unnest_tokens(word, text) %>%
    
    # Remove stopwords
    anti_join(stop_words) %>%
    
    # Count
    count(line, word) %>%
    
    # Calculate and bind the term frequency and inverse document frequency
    bind_tf_idf(word, line, n)

tidy_sms_tbl
```

```{r}
wide_sms_tbl <- tidy_sms_tbl %>%
    select(line, word, tf_idf) %>%
    pivot_wider(
        names_from   = word,
        names_prefix = "word_",
        values_from  = tf_idf,
        values_fill  = 0
    )

wide_sms_tbl
```

## Visualizing text data - word clouds

This part inspired by Julia Silge & David Robinson's book Text Mining with R: A Tidy Approach: https://www.tidytextmining.com/

### Count word frequencies

```{r}
frequency_tbl <- sms_tbl %>%
    
    mutate(
        
        # Make everything lowercase
        text = str_to_lower(text),
        
        # Remove numbers and punctuation,
        text = str_replace_all(text, "[[:punct:][:digit:]]+", " ")
        
    ) %>%
    
    # One word per one row
    unnest_tokens(word, text) %>%
    
    # Removing stop words
    anti_join(stop_words) %>%
    
    # Stemming
    mutate(word = wordStem(word)) %>%
    
    # Count the words
    count(.type, word) %>%
    
    # Count the proportion of words
    mutate(
        proportion = n / sum(n),
        .by        = .type
    ) %>%
    
    # Reorder the columns
    select(-n) %>%
    pivot_wider(names_from = .type, values_from = proportion) %>%
    pivot_longer(
        cols      = c("ham", "spam"),
        names_to  = ".type",
        values_to = "freq"
    )

frequency_tbl
```

### Subset the frequency data into two groups, spam and ham

```{r}
spam_tbl <- frequency_tbl %>%
    filter(.type == "spam") %>%
    select(-.type) %>%
    drop_na()

spam_tbl %>%
    arrange(desc(freq))
```

```{r}
ham_tbl <- frequency_tbl %>%
    filter(.type == "ham") %>%
    select(-.type) %>%
    drop_na()

ham_tbl %>%
    arrange(desc(freq))
```

### Word cloud

This part inspired by a blog post by Céline Van den Rul:
https://towardsdatascience.com/create-a-word-cloud-with-r-bde3e7422e8a

```{r}
# One for ham...
ham_tbl %>%
    wordcloud2(
        size            = 2,
        color           = "random-light",
        backgroundColor = "black"
    )
```

```{r}
# ...and another for spam
spam_tbl %>% 
    wordcloud2(
        size            = 2,
        color           = "random-light",
        backgroundColor = "black"
    )
```

## Creating the recipe and splitting the data

### Create the recipe

```{r}
text_recipe_obj <- recipe(
    .type ~ text,
    data = sms_tbl_row_numbers
) %>%
    step_tokenize(text)  %>%
    step_stopwords(text) %>%
    step_tokenfilter(text, max_tokens = 1e3) %>%
    step_tfidf(text)

text_recipe_obj
```

```{r}
# Bake it
sms_baked_tbl <- text_recipe_obj %>%
    prep() %>%
    bake(new_data = NULL)

sms_baked_tbl
```

```{r}
# Simplify the tf-idf to yes/no
sms_baked_longer_tbl <- sms_baked_tbl %>%
    mutate(
        across(
            where(is.numeric),
            ~ case_when(
                . > 0 ~ "Yes",
                .default = "No"
            )
        )
    ) %>%
    
    # Rename the columns back to words
    rename_with(
        ~ tolower(str_remove(.x, "tfidf_text_")),
        .cols = -starts_with(".type")
    )

sms_baked_longer_tbl
```

### Create training and test data

Not randomly, because the messages weren't in any particular order

```{r}
sms_split <- initial_time_split(
    sms_baked_longer_tbl,
    prop = 0.75
)

sms_train <- training(sms_split)
sms_test  <- testing(sms_split)
```

## Training a model on the data

naivebayes is the engine (needs to be installed if not already):
install.packages("discrim") AND
install.packages("naivebayes")

It is used as the engine for {parsnip}'s naive_Bayes() function. And since we are classifying, that is the mode we choose.

The simple reason is I couldn't get klaR (the other engine) to work. If you know how, please comment on GitHub. It would be great to get to test what the difference between the two engines are.

### Model specification

```{r}
model_spec <- naive_Bayes(
    engine     = "naivebayes",
    mode       = "classification",
    smoothness = NULL,
    Laplace    = NULL
) %>%
    translate()

model_spec
```

### Fit the model

```{r}
model_fit <- model_spec %>% 
    fit(
        .type ~ .,
        sms_train
    )

model_fit
```

### Make the predictions (you could skip this step)

```{r}
sms_test_pred <- model_fit %>% 
    predict(
        new_data = sms_test,
        type     = "class"
    )

sms_test_pred
```

### Add the predictions to the test tibble

```{r}
sms_test_with_pred_tbl <- augment(model_fit, sms_test)
sms_test_with_pred_tbl
```

## Evaluating model performance

### Create a confusion matrix

```{r}
conf_mat <- conf_mat(
    data     = sms_test_with_pred_tbl,
    truth    = .type,
    estimate = .pred_class
)

conf_mat
```

### Visualize the confusion matrix

```{r}
conf_mat %>%
    autoplot(type = "heatmap")
```

```{r}
conf_mat %>%
    autoplot(type = "mosaic")
```

### Visualize the ROC curve

```{r}
sms_test_with_pred_tbl %>%
    roc_curve(
        truth = .type,
        .pred_ham
    ) %>%
    autoplot()
```

### Calculate the ROC AUC (area under the curve)

```{r}
sms_roc_auc <- sms_test_with_pred_tbl %>%
    roc_auc(
        truth    = .type,
        .pred_ham
    )

sms_roc_auc
```

## Compare to k-NN

### Read the results

```{r}
sms_results_knn <- read_csv("data/sms_results_knn.csv") %>%
    mutate(p_ham = 1 - p_spam)
```

### Naive Bayes vs k-NN - ROC curve

```{r}
# Naive Bayes
sms_test_with_pred_tbl %>%
    roc_curve(
        truth = .type,
        .pred_ham
    ) %>%
    autoplot()
```

```{r}
# k-NN
sms_test_with_pred_tbl %>%
    mutate(.knn_p_ham = sms_results_knn$p_ham) %>%
    roc_curve(
        truth = .type,
        .knn_p_ham
    ) %>%
    autoplot()
```

### Naive Bayes vs k-NN - AUC

```{r}
# Naive Bayes
sms_test_with_pred_tbl %>%
    roc_auc(
        truth    = .type,
        .pred_ham
    )
```

```{r}
# k-NN
sms_test_with_pred_tbl %>%
    mutate(.knn_p_ham = sms_results_knn$p_ham) %>%
    roc_auc(
        truth    = .type,
        .knn_p_ham
    )
```

### Put together other model metrics

```{r}
# Such as accuracy, Matthews correlation coefficient (mcc) and others...
classification_metrics <- conf_mat(
    sms_test_with_pred_tbl,
    truth    = .type,
    estimate = .pred_class
) %>%
    summary()

classification_metrics
```

## Improving model performance

Basically, the same as before, but with Laplace = 1

### Model specification

```{r}
model_spec <- naive_Bayes(
    engine     = "naivebayes",
    mode       = "classification",
    smoothness = NULL,
    Laplace    = 1
) %>%
    translate()

model_spec
```

### Fit the model

```{r}
model_fit <- model_spec %>%
    fit(
        .type ~ .,
        sms_train
    )

model_fit
```

### Make the predictions (you could skip this step)

```{r}
sms_test_pred <- model_fit %>% 
    predict(
        new_data = sms_test,
        type     = "class"
    )

sms_test_pred
```

### Add the predictions to the test tibble

```{r}
sms_test_with_pred_tbl <- augment(model_fit, sms_test)
sms_test_with_pred_tbl
```

## Evaluating model performance

### Create a confusion matrix

```{r}
conf_mat <- conf_mat(
    data     = sms_test_with_pred_tbl,
    truth    = .type,
    estimate = .pred_class
)

conf_mat
```

### Visualize the confusion matrix

```{r}
conf_mat %>% autoplot(type = "heatmap")
```

```{r}
conf_mat %>% autoplot(type = "mosaic")
```

### Visualize the ROC curve

```{r}
sms_test_with_pred_tbl %>%
    roc_curve(
        truth = .type,
        .pred_ham
    ) %>%
    autoplot()
```

### Calculate the ROC AUC (area under the curve)

```{r}
sms_roc_auc <- sms_test_with_pred_tbl %>%
    roc_auc(
        truth = .type,
        .pred_ham
    )

sms_roc_auc
```

### Put together other model metrics

Such as accuracy, Matthews correlation coefficient (mcc) and others...

```{r}
classification_metrics <- conf_mat(
    sms_test_with_pred_tbl,
    truth    = .type,
    estimate = .pred_class
) %>%
    summary()

classification_metrics
```

## Creating a function to help evaluate the model further

The assumption here is that you have already gone through steps 1. to 4. What we're potentially tuning here are the arguments .smoothness and .Laplace. Check out the book and/or the documentation for further info about them!

```{r}
classify_with_naive_bayes <- function(
        .smoothness  = NULL,
        .laplace     = NULL
) {
    
    # Model specification
    model_spec <- naive_Bayes(
        engine     = "naivebayes",
        mode       = "classification",
        smoothness = .smoothness,
        Laplace    = .laplace
    ) %>%
        translate()
    
    # Fit the model
    model_fit <- model_spec %>% 
        fit(
            .type ~ .,
            sms_train
        )
    
    # Add the predictions to the test tibble
    sms_test_with_pred_tbl <- augment(model_fit, sms_test)
    
    # Create a confusion matrix
    conf_mat <- conf_mat(
        data     = sms_test_with_pred_tbl,
        truth    = .type,
        estimate = .pred_class
    )
    
    # Print the confusion matrix
    conf_mat %>% autoplot(type = "heatmap")
    
}
```

### Test the function

```{r}
classify_with_naive_bayes(
    .smoothness  = 1,
    .laplace     = 1
)
```

## Understanding the classifier's predictions

### Obtain the predicted probabilities (you could skip this step)

```{r}
sms_test_prob <- model_fit %>% 
    predict(
        new_data = sms_test,
        type     = "prob"
    )

sms_test_prob
```

### Look at the predicted probabilities

```{r}
sms_results <- sms_test_with_pred_tbl %>%
    select(
        actual_type    = .type,
        predicted_type = .pred_class,
        prob_spam      = .pred_spam,
        prob_ham       = .pred_ham
    ) %>%
    mutate(
        prob_spam = prob_spam %>% round(2),
        prob_ham  = prob_ham %>% round(2)
    )

sms_results
```

### Test cases where the model is less confident

```{r}
sms_results %>%
    filter(between(prob_spam, 0.40, 0.60))
```

### Test cases where the model was wrong

```{r}
sms_results %>%
    filter(actual_type != predicted_type)
```

### Specifying vectors

```{r}
sms_results %>%
    tabyl(actual_type, predicted_type)
```

### Using {crosstable}

```{r}
sms_results %>%
    crosstable(
        cols            = predicted_type,
        by              = actual_type,
        label           = FALSE,
        total           = "both",
        percent_pattern = "{n} ({p_row}/{p_col}/{p_tot})",
        percent_digits  = 1
    ) %>%
    as_flextable(compact = TRUE)
```
